---
title: "STAT/MATH 495: Problem Set 06"
author: "Syed Abbas Shah"
date: "2017-10-17"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    collapsed: false
    smooth_scroll: false
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, fig.width=8, fig.height=4.5, message=FALSE, warning = FALSE
  )
set.seed(76)

# Load packages
library(tidyverse)
library(mosaic)
library(broom)
library(knitr)
library(gridExtra)
```





# Collaboration

Please indicate who you collaborated with on this assignment: 

No one



# Setup

Define truth, which again we know for the purposes of this assignment, but in
practice we won't:

* the true function f(x) i.e. the signal
* the true epsilon i.e. the noise, which in this case is Normal$(0, sd=\sigma)$.
Hence the standard deviation $\sigma$ determines the amount of noise.

```{r}
f <- function(x) {
  x^2
}
sigma <- 0.3
```

This is the target point we'll be trying to predict: $(0.95, f(0.95)) = (0.95, 0.95^2) = (0.95, 0.9025)$, Thus, the test set is just `x=0.95`

```{r}
x0 <- 0.95
test_set <- data_frame(x=x0)
```

This function generates a random sample of size $n$; think of this as a "get new
data" function. Random in terms of both:

* (New) the predictor x (uniform on [0,1])
* the amount of noise $\epsilon$

```{r}
generate_sample <- function(f, n, sigma) {
  sample <- data_frame(
    x = runif(n = n, min = 0, max = 1),
    f_x = f(x),
    epsilon = rnorm(n = n, mean = 0, sd = sigma),
    y = f_x + epsilon
  )
  # Recall: We don't observe f(x) and epsilon, just (x, y)
  sample <- sample %>% 
    select(x, y)
  return(sample)
}
```

Define

* The number $n$ of observations $(x_i, y_i)$ in each sample. In the handout,
$n=100$ to keep plots uncrowded. Here we boost to $n=500$
* Number of samples of size $n$ to consider

```{r}
n <- 500
n_sample <- 10000
```


# Computation


```{r, echo=FALSE, message=FALSE, warning=FALSE}
s<-generate_sample(f,n,sigma)
```



```{r, echo=FALSE, message=FALSE, warning=FALSE}
baseplot <- ggplot(data=NULL) +
  geom_vline(xintercept = x0, linetype="dashed") +
  geom_point(data=test_set, aes(x=.95,y=.9025), col="brown", size=4.5) +
  coord_cartesian(xlim=c(0, 1), ylim=c(-0.75, 1.5))

plot_df_2 <- baseplot +
  labs(title="Degrees of freedom = 2")
plot_df_99 <- baseplot +
  labs(title="Degrees of freedom = 99")
```


Let's try our two spline models on 40 samples randomly generated from our function. This is so we can visualize what we're trying to do here.

```{r, message=FALSE, warning=FALSE}
for(i in 1:40) {
  sampled_points <- generate_sample(f, n, sigma)
  # Fit splines with df=2 and add to plot
  fitted_df_2 <- smooth.spline(x=sampled_points$x, y=sampled_points$y, df=2) %>%
    augment()
  plot_df_2 <- plot_df_2 +
    geom_point(data=fitted_df_2, aes(x=x, y=y), alpha=0.1) +
    geom_line(data=fitted_df_2, aes(x=x, y=.fitted), col="blue", alpha=.8, size=1.25)
  # Fit splines with df=99 and add to plot
  fitted_df_99 <- smooth.spline(x=sampled_points$x, y=sampled_points$y, df=99) %>%
    augment()
  plot_df_99 <- plot_df_99 +
    geom_point(data=fitted_df_99, aes(x=x, y=y), alpha=0.1) +
    geom_line(data=fitted_df_99, aes(x=x, y=.fitted), col="blue", alpha=0.8, size=1.25)
  
}
plot_df_2 <- plot_df_2 +
  geom_point(data=data_frame(x=x0, y=0.9025), aes(x=x,y=y), col="red", size=4.5)
plot_df_99 <- plot_df_99 +
  geom_point(data=data_frame(x=x0, y=0.9025), aes(x=x,y=y), col="red", size=4.5)
grid.arrange(plot_df_2,plot_df_99)
```


# Tables

Let's simulate 500 points and compute values for the error terms

```{r, message=FALSE, warning=FALSE}
x0<-0.95
y0 <- .95^2 #actual y0
y<-0
z<-0
set.seed(50)
for(i in 1:500){  
sampled_points <- generate_sample(f, n, sigma)
model2 <- smooth.spline(x=sampled_points$x, y=sampled_points$y, df=2)
y[i]<- predict(model2,x0)$y
model3 <- smooth.spline(x=sampled_points$x, y=sampled_points$y, df=99)
z[i]<- predict(model3,x0)$y
}  
y<- data.frame(y)
resid <- y0 - y$y
resid2 <- resid^2
var_y <- var(y$y)
mse <- mean (resid2)
bias <- mean(y$y) - y0
bias2 <- bias^2
sdeps <- sd(resid)
vareps <- var(resid)
bias2 + mse + sdeps

z<- data.frame(z)
resid <- y0 - z$z
resid2 <- resid^2
var_z <- var(z$z)
mse <- mean (resid2)
bias <- mean(z$z) - y0
bias2 <- bias^2
sdeps <- mean(resid)
var(resid)
bias2 + mse 

```


I will record the responses here:


 Df 2:
 The MSE is 0.146
 Bias squared is .0139
 Var is 0.0006
 
 for Df 99:
 MSE is 0.0189
 bias squared is 4.36e-06
 Var is 0.0189









# Analysis

**Questions**:

1. Based on the topics covered in Lec 2.7, name one possible "sanity check" for your results. Name another if you can.

2. In **two** sentences or less, give a rough sketch of what the procedure would
be to get the breakdown of $$\mbox{MSE}\left[\widehat{f}(x)\right]$$ for *all*
$x$ in this example, and not just for $$\mbox{MSE}\left[\widehat{f}(x_0)\right]
= \mbox{MSE}\left[\widehat{f}(0.95)\right]$$.

3. Which of the two models would you choose for predicting the point of interest and why?

**Answers**:

1. The plots should show that low flexibility models should be have lower variance (fewer 'wiggles') as opposed to higher flexibility models. Moreover, the overeall error can never be lower than the irreducible error. If that appears to be the case, something has gone wrong.

2. We would have to look at the bias and variance in estimates for all values by computing the difference between the estimates and the true values. 


3. If the goal were simply to predict the point of interest, I would for the more flexible model i.e. the df=99 one. This is because the bias of this model is much lower than the lm model (even if the variance is higher). As we can see in the plots above, the flexible curves contain the red dot, whereas the linear lines tend to miss it.

